The Geometry of Archaeological Thought: Analysing discplinary change using word embeddings from conference abstracts

### 2020-05-07 documents/paper.Rmd
stash

The study of disciplinary change traditionally relies on close reading of a large number of publications, manually sorting them into pre-defined categories and making connections between these groups and external influences [e.g. @trigger1989history]. These methods are time-consuming and expensive, which excludes many people from participating. They are also vulnerable to biases

computational approaches that could scan the entire corpus to generate topics in a bottom-up fashion, based on the semantic content of the publications themselves. These approaches indeed make it possible to analyze more comprehensively and more systematically the contents of large corpora of full-text documents, be it in history and sociology

@@

@@

- source of the data
We collected abstracts from Society of American Archaeology (SAA) annual meeting programs from 1964 to 2020. The total number of abstracts are 68,176 from 40 annual meetings excluding some years without available abstracts. The source of abstract includes papers, posters, workshop posters, symposium represented by paragraphs. The annual programs are available online at https://www.saa.org/annual-meeting/annual-meeting-archives/program-archives. The original format of abstracts on websites includes non-searchable PDF documents coming from images for early years and searchable PDF documents after 2005.

- preparation steps: data cleaning (Cite the quanteda pkg, and things they cite)
For image-based pdfs, we convert them into PNG files, clean up the images using Magick package, and then extract the text using Tesseract package [@Kay2007] for optical character recognition (OCR). We obtain 95-98 percent accuracy of OCR on a word level across years according to the results of error rate for each year. After the text are ready, we use quanteda package for text managing and analyzing [@Benoit2018]. We clean up the text by removing all punctuation and stop words, such as "the", "an", "a", that might have some impacts on our results.

- word embeddings (cite our readings)
Word embeddings is useful tool based machine learning framework for measurement, quantifying, and comparing the semantic relations between words, which enables a wide range of application for temporal changes in word relationships related to the sociocultural domain [@Garg2018; @Kozlowski2019]. There are two major word-level pretrained word embeddings, Weod2Vec and Glove according to different methods [@Rong2014; @Pennington2014]. Since pretrained word embeddings are trained on large corpus such as Google News, they can capture the semantic meaning  of a word very well.

In this article, we use text2vec package [@Selivanov2016] to train the word embeddings using GloVe model, which calculates the occurrence of a pair of words and creates the co-occurrence matrix.

- temporal referencing plots (cite our readings)

>>>>

The study of disciplinary change traditionally relies on close reading of a large number of publications, manually sorting them into pre-defined categories and making connections between these groups and external influences [e.g. @trigger1989history]. These methods are time-consuming and expensive (because they depend on full access to a large library of texts), which limits participation in this aspect of archaeology. Computational approaches to analyzing text that can process many publications quickly to generate topics in a bottom-up fashion offer a new way to investigate disciplinary history. This has the advantage of increasing and diversifying participation in study of the history of archaeology, as well as  more comprehensive, systematic, and reproducible analyses of the contents of large number of full-text documents. The methods and data we present here offer empirical basis for what might otherwise be informal claims about the discipline and its history. They are also the type of data that may prompt discussion about the directions the discipline might — or could — take.

In this paper we present a dataset of all available abstracts of papers and posters presented at annual meetings of the Society of American Archaeology. We demonstrate the use of word embeddings, a commonly used computational tool in natural language processing (NLP) and machine learning, as a method to measure, quantify, and compare changes in the way archaeologists write. We present the temporal reference plot as a method for visualizing disciplinary change in archaeology. As a case study, we use these data and methods to explore how archaeologists' writing about explaining and theorizing the past have changed over time.

@@

Research into the history of archaeology is often organised into two major phases. An early internalist phase, where the story of development of the field is told with limited reference to the influence of political, social, religious, and economic institutions. In the 1980s an external, critical history phase began, seeking to understand the interactions between archaeological practice and its political, economic and social contexts.

Among historians of archaeology there are two major phases in  ‘critical history of archaeology’ in the 1980s had been characterized by the replacement of an

@@

- source of the data
We collected abstracts from Society of American Archaeology (SAA) annual meeting programs from 1964 to 2020. The total number of abstracts are 68,176 from 40 annual meetings excluding some years without available abstracts. The source of abstract includes papers, posters, workshop posters, symposium represented by paragraphs. The annual programs are available online at https://www.saa.org/annual-meeting/annual-meeting-archives/program-archives. The original format of abstracts on websites includes scanned images for early years and accessible PDF after 2004.

- preparation steps: data cleaning (Cite the quanteda pkg, and things they cite)

- word embeddings (cite our readings)
- temporal referencing plots (cite our readings)
